Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-07-27 manuscript/usses_water_ms.Rmd
more discussion writing

<!---
NOTES FROM MEETING WITH PETER
1. Bet hedgers -- dominant species here deal well with low water and interannual variability; they live in dry and variable in environment, which has selected for tratis (not hitting physiological thresholds which only becom apparaent when species really go for it). Bet hedging = do the same things every year.
2. Statistical explanation -- either (1) above explains our weak and variable relationship, or we just looked at the wrong thing; lots of other things beyond VWC that are influencing ANPP.
3. Limitation -- snow, we don't know what happens. Its blowing around, who knows how big of an impact this has.
4. Show Fig 2D with real model fits and move to results figure 3.
---->

@@

The similarity of ecosystem functional response (Fig. 3A) and community composition (Fig. 5) across treatments is surprising because grasslands generally, and sagebrush steppe specifically, are considered water-limited systems.
Indeed, we expected ecosystem functional response, community composition, or both to change with precipitation treatment, landing us in any box of Fig. 1 \emph{except} the top left.
So, why did our treatments fail to induce ecosystem or community responses?
We can think of three reasons; two are limitations of our study, and one is the life history traits of the species in our focal communities.
We first discuss the potential limitations of our study, and then discuss biological explanations.

@@

Second,

Third, the ecosystem and community resistance to drought and deluge we observed could be a real phenomenon in this system, at least over the five year time span we observed.
The perennial plants in this cold desert ecosystem are tolerant to drought conditions [@Bazzaz1979; @Franks2011, A.R. Kleinhesselink, unpublished data], meaning a consistent ecosystem functional response at low soil moisture is perfectly reasonable.
For example, many of the perennial grasses in our focal ecosystem avoid drought stress by growing early in the growing season [@Blaisdell1958, A.R. Kleinhesselink, personal observation].
Likewise, the dominant shrub in our focal ecosystem, *Artemisia tripartita*, has access to water deep in the soil profile thanks to a deep root system (CITATION).
As for the irrigation treatments, the lack of a saturating response (e.g., a lower regression slope) suggests water availability remained a limiting factor at the levels of precipitation we were able to manipulate.

In conclusion, our results suggest that five years of $\pm$ 50\% ambient precipitation is not enough to induce a shift in ecosystem functional response in a sagebrush steppe.
This is likely because the species in our focal plant community are bet hedgers, maintaining relatively consistent responses to interannual variation in precipitation to avoid booms and busts.
Longer time series of chronic precipitation alteration may reveal plant community shifts that we did not observe [e.g., @Wilcox2016], in which case species that do not bet hedge may gain prominence and dominate the ecosystem functional response.
Our results suggest compositional shifts would have the largest impact at high rainfall because the current community maintained consistent ecosystem functional response at very low water availability.

>>>>

@@

The similarity of ecosystem functional response (Fig. 3A) and community composition (Fig. 5) across treatments is surprising because grasslands generally, and sagebrush steppe specifically, are considered water-limited systems.
Indeed, we expected ecosystem functional response, community composition, or both to change with precipitation treatment, landing us in any box of Fig. 1 \emph{except} the top left.
So, why did our treatments fail to induce ecosystem or community responses?
We can think of three reasons; two are limitations of our study, and one is the life history traits of the species in our focal communities.
We first discuss the potential limitations of our study, and then discuss a biological explanation.

@@

Second, ANPP at our site may be influenced by factors beyond the window of soil moisture we included in our statistical model.
For example, temperature can also impact community composition and ANPP [@Hoover2014].
Measurements of soil moisture likely contain a signal of temperature, through its impact on evaporation and infiltration, but the measurements will not capture the direct effect of temperature on metabolic and physiological processes.
Likewise, we did not redistribute snow across our treatments in the winter, and snowmelt may spur early spring growth (CITATION).

Third, the life history traits of the dominant species in our study ecosystem may explain the consistent and weak effect of soil moisture on ANPP (Fig. 3).
Species that live in variable environments, such as cold deserts, must have strategies to ensure long-term success as conditions vary.
One strategy is bet hedging, where species forego short-term gains to reduce the variance of long-term success (CITATION).
The dry and variable environment of the sagebrush steppe has likely selected for bet hedging species that can maintain function at low water availability and have weak responses to high water availability.
In so doing, the dominant species in our ecosystem avoid "boom and bust" cycles, which corresponds to the weak effect of soil moisture on ANPP (i.e., the credible intervals for $\beta_1$ overlapping zero), even at precipitation extremes.

Another strategy to deal with variable environmental conditions is avoidance (CITATION), which would result in a consistent ecosystem functional response at low soil moisture.
The perennial plants in this cold desert ecosystem are tolerant to drought conditions [@Bazzaz1979; @Franks2011, A.R. Kleinhesselink, unpublished data].
For example, many of the perennial grasses in our focal ecosystem avoid drought stress by growing early in the growing season [@Blaisdell1958, A.R. Kleinhesselink, personal observation].
Likewise, the dominant shrub in our focal ecosystem, *Artemisia tripartita*, has access to water deep in the soil profile thanks to a deep root system (CITATION).

In conclusion, our results suggest that five years of $\pm$ 50\% ambient precipitation is not enough to induce a shift in ecosystem functional response in a sagebrush steppe.
This is likely because the species in our focal plant community are tolerant of drought conditions and bet hedgers in wet conditions, maintaining relatively consistent responses to interannual variation in precipitation to avoid booms and busts.
Longer time series of chronic precipitation alteration may reveal plant community shifts that we did not observe [e.g., @Wilcox2016], in which case species that do not bet hedge may gain prominence and dominate the ecosystem functional response.
Our results suggest compositional shifts would have the largest impact at high rainfall because the current community maintained consistent ecosystem functional response at very low water availability.

### 2017-07-27 manuscript/usses_water_ms.tex
more discussion writing

Second,

Third, the ecosystem and community resistance to drought and deluge we
observed could be a real phenomenon in this system, at least over the
five year time span we observed. The perennial plants in this cold
desert ecosystem are tolerant to drought conditions \citep[A.R.
Kleinhesselink, unpublished data]{Bazzaz1979, Franks2011}, meaning a
consistent ecosystem functional response at low soil moisture is
perfectly reasonable. For example, many of the perennial grasses in our
focal ecosystem avoid drought stress by growing early in the growing
season \citep[A.R. Kleinhesselink, personal observation]{Blaisdell1958}.
Likewise, the dominant shrub in our focal ecosystem, \emph{Artemisia
tripartita}, has access to water deep in the soil profile thanks to a
deep root system (CITATION). As for the irrigation treatments, the lack
of a saturating response (e.g., a lower regression slope) suggests water
availability remained a limiting factor at the levels of precipitation
we were able to manipulate.

In conclusion, our results suggest that five years of \(\pm\) 50\%
ambient precipitation is not enough to induce a shift in ecosystem
functional response in a sagebrush steppe. This is likely because the
species in our focal plant community are bet hedgers, maintaining
relatively consistent responses to interannual variation in
precipitation to avoid booms and busts. Longer time series of chronic
precipitation alteration may reveal plant community shifts that we did
not observe \citep[e.g.,][]{Wilcox2016}, in which case species that do
not bet hedge may gain prominence and dominate the ecosystem functional
response. Our results suggest compositional shifts would have the
largest impact at high rainfall because the current community maintained
consistent ecosystem functional response at very low water availability.

>>>>

Second, ANPP at our site may be influenced by factors beyond the window
of soil moisture we included in our statistical model. For example,
temperature can also impact community composition and ANPP
\citep{Hoover2014}. Measurements of soil moisture likely contain a
signal of temperature, through its impact on evaporation and
infiltration, but the measurements will not capture the direct effect of
temperature on metabolic and physiological processes. Likewise, we did
not redistribute snow across our treatments in the winter, and snowmelt
may spur early spring growth (CITATION).

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistent and weak effect of soil moisture on
ANPP (Fig. 3). Species that live in variable environments, such as cold
deserts, must have strategies to ensure long-term success as conditions
vary. One strategy is bet hedging, where species forego short-term gains
to reduce the variance of long-term success (CITATION). The dry and
variable environment of the sagebrush steppe has likely selected for bet
hedging species that can maintain function at low water availability and
have weak responses to high water availability. In so doing, the
dominant species in our ecosystem avoid ``boom and bust'' cycles, which
corresponds to the weak effect of soil moisture on ANPP (i.e., the
credible intervals for \(\beta_1\) overlapping zero), even at
precipitation extremes.

Another strategy to deal with variable environmental conditions is
avoidance (CITATION), which would result in a consistent ecosystem
functional response at low soil moisture. The perennial plants in this
cold desert ecosystem are tolerant to drought conditions \citep[A.R.
Kleinhesselink, unpublished data]{Bazzaz1979, Franks2011}. For example,
many of the perennial grasses in our focal ecosystem avoid drought
stress by growing early in the growing season \citep[A.R.
Kleinhesselink, personal observation]{Blaisdell1958}. Likewise, the
dominant shrub in our focal ecosystem, \emph{Artemisia tripartita}, has
access to water deep in the soil profile thanks to a deep root system
(CITATION).

In conclusion, our results suggest that five years of \(\pm\) 50\%
ambient precipitation is not enough to induce a shift in ecosystem
functional response in a sagebrush steppe. This is likely because the
species in our focal plant community are tolerant of drought conditions
and bet hedgers in wet conditions, maintaining relatively consistent
responses to interannual variation in precipitation to avoid booms and
busts. Longer time series of chronic precipitation alteration may reveal
plant community shifts that we did not observe
\citep[e.g.,][]{Wilcox2016}, in which case species that do not bet hedge
may gain prominence and dominate the ecosystem functional response. Our
results suggest compositional shifts would have the largest impact at
high rainfall because the current community maintained consistent
ecosystem functional response at very low water availability.
